COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-06-22 SC-paper/bare_conf.tex
Started re-writing section background

\section{Introduction}
In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by partial differential equations (PDEs). Numerical techniques, such as the finite volume method and finite element method, are widely employed to approximate solutions of these PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each element of the discretized domain by applying suitable numerical operations. As the number of elements can be of the order of trillions \cite{Rossinelli2013}, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful port of Hydra, an industrial computational fluid dynamics application developed by Rolls Royce for turbomachinery design (based on finite volume method, roughly 50,000 lines of code and with meshes in excess of 100M edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach in the implementation of PDE solvers~\cite{IstvanHydra}.

@@

\section{Background}
\label{sec:background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.

\begin{algorithm}[t]
\footnotesize
\KwIn{element matrix (2D array, initialized to 0),\\~~~~~~~~~element coordinates (array),\\~~~~~~~~~coefficient fields (array, e.g. velocity)}
\KwOut{element matrix (2D array)}
\nlset{1}~~- Compute Jacobian from coordinates \\
~~- Declare of constant data: basis functions and derivatives\\
~~- Compute element matrix using numerical quadrature
\caption{General structure of a local assembly kernel generated by Firedrake.}
\label{code:general-structure}
\end{algorithm}

@@

Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution. The process consists of numerically evaluating an integral to produce a small dense element matrix or an element vector~\cite{quadrature1, fluidity_manual_v4}. This operation is applied to all cells in the discretized domain. In this work we focus on local matrices, or ``element matrices'', which are more costly to compute than element vectors.

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. It then triggers compilation of such kernels using a vendor compiler, and manages parallel execution over the mesh. The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating the values of velocity or pressure in the element. The output is the evaluated element matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of a problem-specific set of constant two dimensional arrays of double precision floats containing the values of the basis functions and their derivatives at the integration points; 3) evaluation of the element matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e. using \texttt{static const} in C). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel objects.

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{C|C|C}
\hline
Object name & Type & Variable name(s) \\\hline
Determinant of the Jacobian matrix & double & det  \\ 
Inverse of the Jacobian matrix & double[] & K \\ 
Coordinates & double** & coords\\ 
Fields & double** & w \\ 
Numerical integration weights & double[] & W \\ 
Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ 
Element matrix & double[][] & A\\ \hline
\end{tabulary}
\end{center}
\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
\label{table:map-name-letters}
\end{table}

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse. In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, in which the numerical integration takes place. In rare cases, as in problems involving hyperelasticity, however, such optimizations might not be sufficient to achieve notable performance improvements. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the number of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so large (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). A comprehensive study of such cases is left as further work.

As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging. It is not always the same set of optimizations that bring performance closest to the machine peak. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.

>>>>

\section{Introduction}
In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by partial differential equations (PDEs). Numerical techniques, such as the finite volume method and finite element method, are widely employed to approximate solutions of these PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each cell, or ``element'', of the discretized domain by applying suitable numerical operations. As the number of elements can be of the order of trillions \cite{Rossinelli2013}, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful port of Hydra, an industrial computational fluid dynamics application developed by Rolls Royce for turbomachinery design (based on finite volume method, roughly 50,000 lines of code and with meshes in excess of 100M edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach in the implementation of PDE solvers~\cite{IstvanHydra}.

@@

\section{Background and Motivating Examples}
\label{sec:background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.

\begin{algorithm}[t]
\footnotesize
\KwIn{element matrix (2D array, initialized to 0),\\~~~~~~~~~element coordinates (array),\\~~~~~~~~~coefficient fields (array, e.g. velocity)}
\KwOut{element matrix (2D array)}
\nlset{1}~~- Compute Jacobian from coordinates \\
~~- Define basis functions \\
~~- Compute element matrix in an affine loop nest
\caption{General structure of a local assembly kernel generated by Firedrake.}
\label{code:general-structure}
\end{algorithm}

@@

Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution. The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1, fluidity_manual_v4}), whose sizes depend on the order of the method. This operation is applied to all cells in the discretized domain. In this work we focus on local matrices, or ``element matrices'', which are more costly to compute than element vectors.

Given a finite element description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl} (UFL), Firedrake generates a C-code kernel implementing assembly using a numerical integration algorithm. The same kernel can be applied to any elements: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates. Firedrake triggers the compilation of an assembly kernel using an available vendor compiler, and manages its (parallel) execution over all elements in the mesh. As already stated, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.

%The code transformations described in this work are also generalizable to non-Firedrake assembly kernels, provided that the same numerical integration algorithm is employed.

The structure of a local assembly kernel is shown in Figure~\ref{code:general-structure}. The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating the values of velocity or pressure in the element. The output is the evaluated element matrix. The kernel body can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse: required for the aforementioned change of coordinates, from the reference element to the one being computed; 2) definition of basis functions: used, intuitively, to interpolate the contribution to the PDE solution over the element. The choice of the basis functions is expressed in UFL directly by users. In the generated code, they are represented as global read-only two dimensional arrays (i.e. using \texttt{static const} in C) of double precision floats; 3) evaluation of the element matrix in an affine loop nest, in which the integration is performed. Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel objects.

%JUNK SENTENCES: Since only the values assumed at the integration points are needed to compute the element matrix, a basis function is implemented as a two dimensional array of double precision floats. The same basis functions can be used to evaluate all element matrices, provided that a suitable change of coordinates from a ``reference'' element to a real one is performed. Therefore, basis functions are declared in the kernel body as global read-only arrays (i.e. using \texttt{static const} in C), rather than passed as function parameters. Basis functions are chosen at the UFL layer based on the problem being solved, so they are actually an input to the code generator.

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{C|C|C}
\hline
Object name & Type & Variable name(s) \\\hline
Determinant of the Jacobian matrix & double & det  \\ 
Inverse of the Jacobian matrix & double & K1, K2, ... \\ 
Coordinates & double** & coords\\ 
Fields & double** & w \\ 
Numerical integration weights & double[] & W \\ 
Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ 
Element matrix & double[][] & A\\ \hline
\end{tabulary}
\end{center}
\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
\label{table:map-name-letters}
\end{table}

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse. In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure. The Helmholtz and Burgers equations, along with the Diffusion problem,

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging. It is not always the same set of optimizations that bring performance closest to the machine peak. For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces. Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger. In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.
